\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis

### 2024-06-06 src/tex/ms.tex
reply updates

Again, performing the same benchmark as above on an NVIDIA A100 with 40GB of memory, we find that waveform evaluations are over an of magnitude faster than serial CPU evaluation.
Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
Note that jit for both CPU and GPU evaluation adds an additional overhead to the first evaluation of the waveform which is not included in the above benchmarks. 
Note also that the GPU benchmark does not include the time taken to move the evaluated waveform from the GPU to the CPU.

@@

One of the primary aims of this paper is to show that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension added, $D$, one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD~\citep{2018arXiv181105031M}, one only needs two function calls to evaluate the derivative of all input parameters, regardless of the number of dimensions.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
In gravitational wave physics one typically wants to compute derivatives of a scalar quantity (e.g., the likelihood) with respect to the intrinsic parameters of the binary (e.g., the masses and spins).
Reverse mode AD is therefore ideal for this task.
For a review of AD methods see~\cite{2018arXiv181105031M}.
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than numerical stability.
However, this might change for waveforms in beyond-GR models, waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases many more parameters can be added.

@@

% The coefficients of the original IMRPhenomD implementation are tuned in subsets of parameters instead of all together. 
Since the coefficients for each segment are tuned individually, the correlation between different parameters are ignored, meaning that the provided best-fit solution may not be the global optimum. 
We therefore aim to improve of the accuracy of the waveform by jointly fitting all coefficients at once.

@@

For training, we use the publically available subset (11 waveforms) of the 19 waveforms used in the original IMRPhenomD paper~\citep{Khan:2015jqa}. 
These 11 waveforms are taken from the SXS catalog~\citep{Boyle:2019kee}.
We then run gradient descent, as described above, until the validation loss stops decreasing (the validation is loss is calculated using the NR waveforms discussed below). 
Figure~\ref{fig:loss_compare} shows the relative error (against a test NR waveform; see below) of the original and optimized waveform as a function of dimensionless frequency.
The vertical dashed lines indicate the stitching points for the phase i.e., when the inspiral is joined onto the merger.
We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
In particular, the relative error in the merger region (in between the two vertical dashed lines) is decreased by half while other regions also show good improvement in accuracy.

In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for a set of test waveforms.
In particular, we use 536 waveforms from the SXS catalog (we choose all waveforms with aligned spins, $\chi_{x,y}<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$)~\citep{Boyle:2019kee}.
One can see that the distribution of mismatches after optimization is generally shifted to lower mismatch compared to the original waveform.
In particular, the peak of the original waveform distribution has moved by nearly an order of magnitude, indicating that our AD-assisted optimization procedure provides an improved implementation of the model.

@@

\begin{figure}[t!]
    \script{chirp_mass_error.py}
    \centering
    \includegraphics[width=\linewidth]{figures/chirp_mass_error.pdf}
    \caption{
        Distribution of Fisher information chirp mass error for a population of nearby binaries.
        Automatic differentiation (AD) was used to compute the derivatives in Eq.~\eqref{eq:fisher}.
        As emphasized in the text, after JIT compilation, each error calculation was between three and five times faster than a similar forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
    }
    \label{fig:chirp_mass_error}
\end{figure}

Forecasting the sensitivity of future experiments is a routine task in GW science. 
Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism~\citep{Cornish:2010kf} is commonly deployed to estimate how well a binary system's parameters could be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood~\citep{Vallisneri:2007ev}.
Although in practice this assumption is often violated for realistic detector noise, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.

@@

The Fisher information matrix for a single detector is typically given by 
\begin{equation}
    \label{eq:fisher}
    \mathcal{I}^{k}_{ij} = (\partial_i h^k | \partial_j h^k) \, ,
\end{equation}
where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h^k$ is the strain measured by the detector which is given by,
\begin{equation}
    h^k(\theta) = F^k_+(\phi) h_{+}(\Xi) + F^k_\times(\phi) h_{\times}(\Xi) \, .
\end{equation}
Note that here we have separated out the extrinsic ($\phi$) and intrinsic ($\Xi$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
\begin{equation}
    \mathcal{I}_{ij} =  \mathcal{I}^{\mathrm{Hanford}}_{ij} + \mathcal{I}^{\mathrm{Livingston}}_{ij} + \mathcal{I}^{\mathrm{Virgo}}_{ij}   \, .
\end{equation}
Finally, we invert the Fisher matrix to calculate the covariance matrix,
which provides forecasted measurement errors and parameter covariances for a signal with parameters $\theta$ observed by the given detector network in the high signal-to-noise limit.

@@

The distribution of chirp mass errors from a population of $5 \times 10^3$ binaries can be seen in Fig.~\ref{fig:chirp_mass_error}.
We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within ~30\%.
This demonstrates that AD can be used to accurately do Fisher forecasting for a large population of events.

@@

After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each sources parameter's - so called PE.
To do this, one typically uses an MCMC or nested sampler~\citep{Skilling2004, multinest, dynesty}.
Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$ hours, while BNS systems can take up to weeks.
Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.,~\texttt{BAYESTAR}][]{Singer:2015ema}. 
Moreover, a number of methods have been developed to speed up PE well below the numbers quoted above~\citep{Dax:2021tsq, Islam:2022afg,Roulet:2022kot, Zackay:2018qdy, Cornish:2021lje, Canizares:2013ywa, Leslie:2021ssu}.
Nevertheless, these do not present the whole picture; fast, general PE therefore remains a key aim of GW data analysis.

@@

Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper}.
Differentiable waveforms therefore represent a useful advancement towards efficient GW science.

>>>>

Again, performing the same benchmark as above on an NVIDIA A100 with 40GB of memory, we find that waveform evaluations are over an order of magnitude faster than serial CPU evaluation.
Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.
Note that jit for both CPU and GPU evaluation adds an additional overhead to the first evaluation of the waveform which is not included in the above benchmarks. 
Note also that the GPU benchmark does not include the time taken to move the evaluated waveform from the GPU to the CPU.

@@

One of the primary aims of this paper is to show that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD is significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension added, $D$, one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD~\citep{2018arXiv181105031M}, one only needs two function calls to evaluate the derivative of all input parameters, regardless of the number of dimensions.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
In gravitational wave physics one typically wants to compute derivatives of a scalar quantity (e.g., the likelihood) with respect to the intrinsic parameters of the binary (e.g., the masses and spins).
Reverse mode AD is therefore ideal for this task.
For a review of AD methods see~\cite{2018arXiv181105031M}.
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than numerical stability.
However, this might change for waveforms in beyond-GR models, waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases many more parameters can be added.

@@

% The coefficients of the original IMRPhenomD implementation are tuned in subsets of parameters instead of all together. 
Since the coefficients for each segment are tuned individually, the correlation between different parameters are ignored, meaning that the provided best-fit solution may not be the global optimum. 
We therefore aim to improve the accuracy of the waveform by jointly fitting all coefficients at once.\footnote{
    For a further breakdown of how the optimization changes across the parameter space, see our accompanying paper~y\citep{Lam:2023oga}.
}

@@

For training, we use the publically available subset (11 waveforms) of the 19 waveforms used in the original IMRPhenomD paper~\citep{Khan:2015jqa}. 
These 11 waveforms are taken from the SXS catalog~\citep{Boyle:2019kee}.
We then run gradient descent, as described above, until the validation loss stops decreasing (the validation loss is calculated using the NR waveforms discussed below). 
Figure~\ref{fig:loss_compare} shows the relative error (against a test NR waveform; see below) of the original and optimized waveform as a function of dimensionless frequency.
The vertical dashed lines indicate the stitching points for the phase i.e., when the inspiral is joined onto the merger.
We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
In particular, the relative error in the merger region (in between the two vertical dashed lines) is decreased by half while other regions also show good improvement in accuracy.

In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches for a set of test waveforms.
In particular, we use 536 waveforms from the SXS catalog i.e., all waveforms with aligned spins, $|\chi_{x,y}|<5\times 10^{-3}$, and eccentricity $<2\times 10^{-3}$~\citep{Boyle:2019kee}.
One can see that the distribution of mismatches after optimization is generally shifted to lower mismatch compared to the original waveform.
In particular, the median mismatch (shown as vertical dashed lines in Fig.~\ref{fig:mismatch_hist}) is reduced by $\sim50$\%, indicating that our AD-assisted optimization procedure provides an improved implementation of the model.

@@

\begin{figure}[t!]
    \script{chirp_mass_error.py}
    \centering
    \includegraphics[width=\linewidth]{figures/chirp_mass_error.pdf}
    \caption{
        Distribution of Fisher information chirp mass error for a population of nearby binaries.
        Automatic differentiation (AD) was used to compute the derivatives in Eq.~\eqref{eq:fisher}.
        As emphasized in the text, after JIT compilation, each error calculation was between three and five times faster than a numerical derivative forecasting code \texttt{GWbench}~\citep{Borhanian:2020ypi}.
    }
    \label{fig:chirp_mass_error}
\end{figure}

Forecasting the sensitivity of future experiments is a routine task in GW science. 
Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism (see e.g.,~\citep{Cornish:2010kf}) is commonly deployed to estimate how well a binary system's parameters could be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood~\citep{Vallisneri:2007ev}.
Although in practice this assumption is often violated for realistic detector noise, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.

@@

The Fisher information matrix for a single detector is typically given by 
\begin{equation}
    \label{eq:fisher}
    \mathcal{I}^{k}_{ij} = (\partial_i h^k | \partial_j h^k) \, ,
\end{equation}
where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h^k$ is the strain measured by the detector which is given by,
\begin{equation}
    h^k(\theta) = F^k_+(\phi) h_{+}(\Xi) + F^k_\times(\phi) h_{\times}(\Xi) \, .
\end{equation}
Note that here we have separated out the extrinsic ($\phi$) and intrinsic ($\Xi$) variables as well as introduced the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
\begin{equation}
    \mathcal{I}_{ij} =  \mathcal{I}^{\mathrm{Hanford}}_{ij} + \mathcal{I}^{\mathrm{Livingston}}_{ij} + \mathcal{I}^{\mathrm{Virgo}}_{ij}   \, .
\end{equation}
Finally, we invert the Fisher matrix to calculate the covariance matrix,
which provides forecasted measurement errors and parameter covariances for a signal with parameters $\theta$ observed by the given detector network in the high signal-to-noise limit.

@@

The distribution of chirp mass errors from a population of $5 \times 10^3$ binaries can be seen in Fig.~\ref{fig:chirp_mass_error}.
We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within ~30\%.\footnote{
    Note that since Ref.~\citep{Borhanian:2020ypi} is based on numerical derivatives, the agreement is not expected to be close to machine precision.
}
This demonstrates that AD can be used to accurately do Fisher forecasting for a large population of events.

@@

After the search algorithms have constructed a list of confidently detected binaries, the next step is to sample from the posterior of each source parameter - so called PE.
To do this, one typically uses an MCMC or nested sampler~\citep{Skilling2004, multinest, dynesty}.
Although robust, both MCMC and nested sampling are slow to converge and are known to perform poorly in high dimensional parameter spaces.
For example, sampling the 15 dimensional parameter space for a BBH system can take $\mathcal{O}(10)$ hours, while BNS systems can take up to weeks.
Dedicated fast samplers have been designed to get approximate posteriors on the sky localization to facilitate follow-up electromagnetic observations \citep[e.g.,~\texttt{BAYESTAR}][]{Singer:2015ema}. 
Moreover, a number of methods have been developed to speed up PE well below the numbers quoted above~\citep{Dax:2021tsq, Islam:2022afg,Roulet:2022kot, Zackay:2018qdy, Cornish:2021lje, Canizares:2013ywa, Leslie:2021ssu}.
Nevertheless, these do not present the whole picture; fast, general PE therefore remains a key aim of GW data analysis.

@@

Although we primarily discuss toy examples in this paper, each can be extended to the full data analysis task, some of which will be shown in upcoming papers~\citep{PEpaper, Lam:2023oga}.
Differentiable waveforms therefore represent a useful advancement towards efficient GW science.
